### 2016-05-16 text/Chapter08.tex
Edits in further work and the rest (Chapter 8)

%===================================================================================================

The main goal of this thesis is to provide a \emph{unified} calculus for tracking context
dependence. We have not achieved this goal yet. In Chapter~\ref{ch:applications}, we identified
two kinds of contextual properties that we further covered separately. Flat coeffects in
Chapter~\ref{ch:flat} track whole-context properties, while structural coeffects, covered in
Chapter~\ref{ch:structural}, track per-variable properties. In this chapter, we unify the two
notions.

Although the results presented in this thesis are mainly of a theoretical nature, we indeed
believe that coeffects should be integrated in main-stream programming languages. This is addressed
in the second part of the chapter, where we outline possible approach for a practical implementation
of coeffect systems. Finally, we conclude the chapter with a brief discussion of an alternative
approach to coeffect system, based on the meta-language style.

\paragraph{Chapter structure and contributions}
\begin{itemize}
\item We start by unifying the two kinds of coeffects discussed so far. In
  Section~\ref{sec:further-unified}, we introduce the \emph{unified coeffect calculus}, which
  generalizes flat and structural systems and can be instantiated to track both flat and structural
  properties.

\item We outline one possible approach for practical implementations of coeffects
  (Section~\ref{sec:further-impl}). The technique is based on embedding comonadic computations
  via a lightweight syntactic extension and using rich type systems for capturing the structure of
  coeffect algebra.

\item Finally, we discuss an alternative approach to defining coeffect systems based on the
  meta-language style, which highlights the relationship between our work and related work arising
  from modal logics such as CMTT \cite{logic-cmtt} (Section~\ref{sec:further-meta}).
\end{itemize}

@@

\begin{itemize}
\item The structural coeffect calculus contains explicit rules for context manipulation
  (weakening, contraction, exchange). In the flat coeffect calculus, these rules are not defined
  explicitly, but are admissible.

@@

\item In the flat coeffect calculus, we distinguish between \emph{splitting} context requirements
  and \emph{merging} of context requirements ($\cpar$ and $\czip$, respectively). In the structural
  coeffect calculus, the operations (which model appending and splitting vectors) are invertible
  and so the structural coeffect algebra uses the same tensor product $\atimes$.
\end{itemize}

\noindent
In the unified calculus presented in this section, we address the three differences as follows.
We include explicit rules for context manipulation in the calculus; in systems that arise from flat
calculi, the structural rules do not change coeffects and can thus be applied freely. We generalize
the structure of coeffect annotations using the notion of a ``container'' which can be specialized
to obtain a single annotation or a vector of annotations. This could potentially be used to capture
other structures such as trees in bunched typing \cite{substruct-bunched}. Finally, we distinguish
between splitting and merging of context requirements (using the notation $\apar$ and $\azip$,
respectively). For structural coeffect calculi, the two operators coincide, but for flat
coeffect calculi, they provide the needed flexibility.

\subsection{Shapes and containers}
Our generalization of coeffect structure using a \emph{coeffect container} is based on containers of
Abbott et al. \cite{types-containers}. Interestingly, containers have later been linked
to comonads by Ahman et al. \cite{comonads-containers}. Intuitively a container describes data
types such as lists, trees or streams. A container is formed by shapes (\eg~lengths of
lists)\footnote{Shapes can also be intuitively thought of as sizes, but this is not fully precise
as some containers may have multiple shapes of the same length, \eg~differently balanced trees.}.
For every shape, we can obtain a set of positions in the container (\eg~offsets in a list of a
specified length). More formally:

@@

\begin{definition}
A \emph{\cclrd{unified coeffect scalar}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set
$\C$ together with elements $\cunit, \czero \in \C$, relation $\cleq$ and binary operations
$\cseq, \cpar$ such that $(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids and
$(\C, \cleq)$ is a pre-order. That is, for all $r,s,t\in \C$:
%
\begin{equation*}
\begin{array}{ccr}
r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t  &
\cunit \;\cseq\; r = r = r \;\cseq\; \cunit &
\textnormal{(monoid)}
\\
r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
\czero \;\cpar\; r = r = r \;\cpar\; \czero &
\textnormal{(monoid)}
\\
\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t&
t\; \cleq\; t &
\textnormal{(pre-order)}
\\
\end{array}
\end{equation*}
%
The following distributivity axioms hold:
\begin{align*}
\quad (\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
\quad \cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
\end{align*}
%
In addition, we require the following two properties of $\czero, \cpar$ and $\cleq$:
\begin{equation*}
\begin{array}{l}
 \quad \cclrd{r}\,\cleq\,\cclrd{r'} \; \Rightarrow \; \forall\cclrd{s}.\;(\cclrd{r}\,\cseq\,\cclrd{s})\;\cleq\;(\cclrd{r'}\,\cseq\,\cclrd{s}) \\
 \quad \czero\;\cleq\;(\czero\,\cseq\,\cclrd{r})
\end{array}
\end{equation*}
\end{definition}

@@

In the coeffect judgment $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$, the coeffect annotation
$\aclrd{\textbf{r}}$ is drawn from the set of coeffect scalars $\C$ indexed by the shape corresponding
to $\Gamma$. We write $\sclrd{s} = \slen{\Gamma}$ for the shape corresponding to $\Gamma$. The operation
$\SP(\sclrd{s})$ returns a \emph{set} of positions and so we can write $\aclrd{\textbf{r}} \in
  \SP(\sclrd{s}) \rightarrow \C$  as a mapping from positions (defined by the shape) to scalar coeffects.
We usually write this as the exponent $\aclrd{\textbf{r}} \in \C^{\SP(\sclrd{s})}$.
The coeffect container is also equipped with an operation that appends shapes (when we concatenate
variable contexts) and two special shapes in $\SH$ representing the empty context and the singleton
context.

@@

Therefore, a coeffect annotation is a \emph{vector}
$\aclrd{\textbf{r}} \in \C^{\SP(\sclrd{n})}$ and assigns a coeffect scalar $\aclrd{\textbf{r}}(i) \in \C$
for each position (corresponding to a variable $x_i$ in the context).
\end{example}

@@

\begin{example}
Similarly to the previous example, we can also define a coeffect container with \emph{no} positions,
\ie~$\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$ where $\SH = \{\sflat \}$, $\SP(\sflat) = \emptyset$,
${|}\Gamma{|}=\sflat$ and $\sflat\,\diamond\,\sflat = \sflat$.
This reduces our system to the simply-typed $\lambda$-calculus with no context annotations, because
$\SP(\sflat)=\emptyset$ and so coeffect annotations would be from the empty set $\C^\emptyset$.
\end{example}

@@

\begin{definition}
Given a \cclrd{unified coeffect scalar} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a
\sclrd{coeffect container} $(\SHP, \slen{-}, \stimes, \sempty, \sunit)$ a \emph{\aclrd{unified
coeffect algebra}} extends the two structures with $(\azip, \apar, \azero)$ where $\azero \in \C^{\SP(\sempty)}$
is a coeffect annotation for the empty context and $\azip, \apar$ are families of operations that
combine coeffect annotations indexed by shapes. That is $\forall \sclrd{n},\sclrd{m} \in \SH$:
%
\begin{equation*}
\apar_{\sclrd{m},\sclrd{n}}, \azip_{\sclrd{m},\sclrd{n}} ~:~ \C^{\SP \sclrd{m}} \times \C^{\SP \sclrd{n}} \rightarrow \C^{\SP (\sclrd{m}\,\stimes\,\sclrd{n})}
\end{equation*}
\end{definition}

\noindent
A coeffect algebra induces the following three additional operations:
%
\begin{equation*}
\begin{array}{rclcrcl}
 \alift{-} &\narrow{:}& \C \rightarrow \C^{\SP(\sunit)}  \\
 \alift{\cclrd{x}} &\narrow{=}& \lambda \_ . \cclrd{x}   \\
\\[-0.5em]
 \aseq_{\sclrd{m}} &\narrow{:}&  \C \times \C^{\SP(\sclrd{m})} \rightarrow \C^{\SP(\sclrd{m})}   \\
 \cclrd{r}\, \aseq\, \aclrd{\textbf{s}} &\narrow{=}& \lambda \sclrd{s} . \cclrd{r}\, \cseq\, \aclrd{\textbf{s}}(\sclrd{s}) \\
\\[-0.5em]
 \alen{-} &\narrow{:}& \C^{\SP(\sclrd{m})} \rightarrow \sclrd{m}\\
 \alen{\aclrd{\textbf{r}}} &\narrow{=}& \sclrd{m}
\end{array}
\end{equation*}
%
The $\alift{-}$ operation lifts a scalar coeffect to a shape-indexed coeffect that is indexed by
the shape of a singleton context. The $\aseq_{\sclrd{m}}$ operation is a left multiplication of a
vector by a scalar. As we always use bold face for vectors and ordinary face for scalars (as well
as a distinct colour), using the same symbol is not ambiguous. We also tend to omit the subscript
$\sclrd{m}$ and write~$\aseq$.

@@

The type system for the unified coeffect calculus is parameterized by a coeffect scalar
$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ together with a coeffect algebra $(\azip, \apar, \azero)$
and the derived constructs $\alift{-}$, $\alen{-}$ and $\aseq$.
As in the structural system, free-variable contexts $\Gamma$ are treated as vectors modulo duplicate
use of variables -- the associativity is built-in. The order of variables matters, but can be changed
using the exchange rule. The context annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$
are shape-indexed coeffects (rather than simple vectors as before). As before, functions are
annotated with coeffects scalars.

@@

Note that we use the terms \emph{merging} and \emph{splitting} in the syntactic (top-down) sense. As
discussed in Chapter~\ref{ch:flat}, Section~\ref{sec:flat-calculus-understanding}, we can also read
the rules in semantic (bottom-up) sense, in which case assumptions merge available contextual
information and conclusions split available contextual information.

@@

%--------------------------------------------------------------------------------------------------

\subsection{Semantics of unified calculus}
\label{sec:further-semantics}

The semantics of the unified calculus can be obtained by generalizing the semantics of the structural
calculus in the same way as we generalized the type system. However, it is worth exploring the type
signatures of the operations of the underlying (comonad-based) structure.

The semantics of the unified coeffect calculus follows the same pattern as the semantics of the
structural calculus (Section~\ref{sec:struct-semantics}) with a number of differences. The
index $\aclrd{\textbf{r}}$ of the object mapping $\ctyp{\aclrd{\textbf{r}}}$ is an element of
the structure $\C^{\SP(\sclrd{n})}$ for a specified container $\SH \triangleleft \SP$ and a shape
$\sclrd{n}$ determined as $\sclrd{n} = \slen{\Gamma}$. To match the type system discussed in
Section~\ref{sec:further-types}, we need to modify the domains and codomains of the \ident{merge},
\ident{split} and \ident{dup} operations as follows.
%
\begin{equation*}
\begin{array}{lcl}
 \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} &\narrow{:}&
  \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \rightarrow \ctyp{\aclrd{\textbf{r}}\azip\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \\
 \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} &\narrow{:}&
  \ctyp{\aclrd{\textbf{r}}\apar\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \rightarrow \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \\
 \ident{dup}_{\cclrd{r}, \cclrd{s}} &\narrow{:}&
  \ctyp{\alift{\cclrd{r} \cpar \cclrd{s}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{r}}\,\azip\,\alift{\cclrd{s}}}{(\alpha \xtimes \alpha)} \\
\end{array}
\end{equation*}
%
The key difference is that the vector concatenation $\atimes$ is replaced with $\apar$ in the
\emph{domain} of the operations (in \ident{split}) and with $\azip$ in the codomain
(in \ident{merge}, but also in \ident{dup} which models contraction).
This means that the splitting and merging the context can change the coeffect annotation:
%
\begin{equation*}
\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} ~:~
  \ctyp{\aclrd{\textbf{r}}\apar\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)}
  \rightarrow \ctyp{\aclrd{\textbf{r}}\azip\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)}
\end{equation*}
%
For flat coeffect calculi where $(\aclrd{r}\,\azip\,\aclrd{s}) \;\aclrd{\leq}\; (\aclrd{r}\,\apar\,\aclrd{s})$,
splitting and then merging a context may lose some of the available contextual capabilities
(and increase context requirements). This illuminates why our proofs of Lemma~\ref{thm:further-rev-contr}
and Lemma~\ref{thm:further-rev-exch} relied on subcoeffecting in the flat system.

\paragraph{Related and future work.}
The semantics presented here serves main\-ly to inform the design of the type system. For this
reason, we keep the semantics simple and close to the concrete notions used in the type system
such as flat, structural and unified coeffect algebras. In particular, the indexed comonad
structure consists of object mapping $\ctyp{r}{}$ that is indexed by a corresponding algebra:

\begin{compactitem}
\item[--] In the flat coeffect calculus, $\ctyp{\cclrd{r}}{}$ is indexed by scalar coeffects $\cclrd{r} \in \C$
\item[--] In the structural coeffect calculus, the object mapping $\ctyp{\aclrd{\textbf{r}}}{}$ is
  indexed by vectors of scalars $\aclrd{\textbf{r}} \in \bigcup_{\sclrd{m}\in \sclrd{\mathbb{N}}} \C^{\sclrd{m}}$
\item[--] In unified coeffect calculus with a coeffect container $\SH\,\sclrd{\triangleleft}\,\SP$,
  the mapping $\ctyp{\aclrd{\textbf{r}}}{}$ is indexed by shape-indexed scalars
  $\aclrd{\textbf{r}} \in \bigcup_{\sclrd{s}\in \SH} \C^{\SP(\sclrd{m})}$
\end{compactitem}

\noindent
In this treatment, the indices are formed by objects that are modelled outside of category theory.
A fully categorical semantics can be found in a joint paper with Orchard \cite{coeffects-icfp14}.
There, the index (coeffect scalars and shape-indexed scalars) are modelled as categories too.
The definition is written using the following notation:

\begin{compactitem}
\item[--] $[\mathbb{C}, \mathbb{D}]$ is a category of functors between categories $\mathbb{C}$ and $\mathbb{D}$
\item[--] $B^A$ is an exponential object $A \Rightarrow B$ in a Cartesian-closed category
\end{compactitem}

\noindent
In the categorical semantics of coeffect calculi, coeffect scalars are modelled as a
category of scalars $\cclrd{\mathbb{I}}$ such that $\textit{obj}(\cclrd{\mathbb{I}}) = \C$.
The structure of the context (indexed comonad) over the input then becomes a functor
indexed by the free-variable context shape $\slen{\Gamma}$ and coeffect annotations
$\aclrd{\textbf{r}}$, \ie~$\ctyp{\slen{\Gamma}}_{\aclrd{\textbf{r}}}$:
%
\begin{equation*}
\sem{\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau} : \ctyp{\slen{\Gamma}}_{\aclrd{R}}
\sem{\Gamma} \rightarrow \sem{\tau}
\end{equation*}
%
The structure $\ctyp{}$ can be thought of as a dependent product of functors $\ctyp{\sclrd{n}}$
over possible shapes $\sclrd{n} \in \SH$ where:
%
\begin{equation*}
\ctyp{\sclrd{n}} : \cclrd{\mathbb{I}}^{\SP(\sclrd{n})} \rightarrow [\mathbb{C}^{\SP(\sclrd{n})}, \mathbb{C}]
\end{equation*}
%
For a fixed context shape $\sclrd{n}$ the functor
$\ctyp{\sclrd{n}} : \cclrd{\mathbb{I}}^{\SP(\sclrd{n})} \rightarrow [\mathbb{C}^{\SP(\sclrd{n})}, \mathbb{C}]$
maps a coeffect indexed by positions $\SP(\sclrd{n})$ to a functor from a context
$\mathbb{C}^{\sclrd{n}}$ to an object in $\mathbb{C}$. That is, given a coeffect annotation
(matching the shape of the context), we get a functor $\in [\mathbb{C}^{\sclrd{n}},\mathbb{C}]$.
From a programming perspective, this functor defines a data structure that models the additional
context provided to the program. The shape of this data structure depends on the coeffect
annotation $\cclrd{\mathbb{I}}^{\sclrd{n}}$. For example, in bounded reuse, the annotation defines
the number of values needed for each variable and the functor will be
formed by lists of length matching the required number.

In the model above, shapes and positions are still treated as ordinary sets. Indeed, in the
joint paper \cite{coeffects-icfp14}, we treat shapes as just sets of sets. In this chapter, we
instead define \emph{coeffect container} as an extension of containers, which have a well-defined
categorical structure \cite{semantics-containers}. A future work is adapting the semantics of
the unified coeffect calculus so that it is based on the semantics of containers. However, note
that context is \emph{not} a (simple) container, because containers are used as indices (coeffect
annotations) of the object that wrap the free-variable context.

% =================================================================================================
%
%      ####                         #       #                   ##
%      #   #                        #                            #
%      #   #  # ##    ###    ###   ####    ##     ###    ###     #
%      ####   ##  #      #  #   #   #       #    #   #      #    #
%      #      #       ####  #       #       #    #       ####    #
%      #      #      #   #  #   #   #  #    #    #   #  #   #    #
%      #      #       ####   ###     ##    ###    ###    ####   ###
%
% =================================================================================================

\section{Towards practical coeffects}
\label{sec:further-impl}

As discussed earlier, the main focus of this thesis is the development of the much needed
\emph{theory of context-aware programming languages} and so discussing the details of a practical
implementation of coeffect tracking is beyond the scope of the thesis. However, this section
briefly outlines one possible pathway towards this goal.

Many of the examples of contextual computation that we discussed earlier have been implemented as
a single-purpose programming language feature (\eg~implicit parameters \cite{app-implicit-parameters}
or distributed computations \cite{app-distributed-ml5,app-distributed-links}). However,
the main contribution of this thesis is that it captures
\emph{multiple} different notions of context-aware computations using just a \emph{single}
common abstraction. For this reason, we advocate that future practical implementations of
coeffects should not be single-purpose language features, but rather reusable abstractions that
can be instantiated with a concrete \emph{coeffect algebra} specified by the user.

In order to do this, programming languages need to provide two features; one that allows
embedding of context-aware computations themselves in programs akin to the ``do'' notation
in Haskell (Section~\ref{sec:further-impl-embed}) and one that allows tracking of the
contextual information in the type system (Section~\ref{sec:further-impl-types}). To make
constraint resolution in type inference easier, we also discuss how the coeffect algebra
can be simplified (Section~\ref{unified-impl-semilattice})

\subsection{Embedding contextual computations}
\label{sec:further-impl-embed}

\subsection{Coeffect annotations as types}
\label{sec:further-impl-types}

The work \cite{effects-embedding} on embedding effect systems in Haskell demonstrated that the
recent additions to the Haskell type system provide enough power to implement structures such
as sets at the type level. Using these to embed coeffect systems in Haskell is one fruitful
future direction for applied coeffects.

In dependently-typed programming languages such as Agda or Idris \cite{other-agda,other-idris},
the embedding of coeffects can be implemented more directly (as terms implementing sets or lattices
can be lifted to the type level). However, we believe that coeffect tracking does not require full
dependent types and can be made accessible in more mainstream languages. Dependent ML
\cite{types-dependent-ml} provides an interesting example of a language with some dependent typing
(\eg~computations with integers) which is still close to its non-dependently-typed predecessor ML.

\subsection{Alternative formulation using coeffect lattice}
\label{sec:further-impl-semilattice}

Compared to typical effect systems based on sets, coeffect systems have richer structure consisting
of four operations ($\cseq,\czip,\cpar$ and a relation $\cleq$). This allows capturing interesting
properties, but it makes working with coeffect annotations difficult -- it complicates tasks
such as type generalization and constraint solving in type inference. Here, we look at one possible
simplification of the structure.

\begin{itemize}
\item As mentioned in Section~\ref{sec:flat-calculus-algebra}, when the monoid $(\C,\cpar,\czero)$ is a
semi-lattice, the $\cleq$ relation can often be expressed in terms of the $\cpar$ operation
($\cclrd{r} \;\cleq\; \cclrd{s} \;\Longleftrightarrow\; \cclrd{r} \;\cpar\; \cclrd{s} \;=\; \cclrd{s}$).
This does not give a definition equivalent to our, but it is a simplification that is consistent
with all examples in this thesis.

\item Another approach that works for most (but not all) systems we discussed is to require the structure
$(\C,\cpar,\czip)$ to form a lattice. This is consistent with all our structural coeffect systems as
well as flat systems for liveness and dataflow, but not with the system for tracking of implicit
parameters (or sets of rebindable resources).
\end{itemize}

\noindent
The system for tracking implicit parameters does not fit the lattice-based model, because it uses the
same operation $\cup$ for both $\cpar$ and $\czip$.

This also means that the system for implicit parameters cannot be used with a system where lambda
abstraction duplicates the context requirements (Section~\ref{sec:flat-exts-lambda}).

\paragraph{Lattice-based coeffects.} Putting implicit parameters aside, there are many other coeffect
systems for which we could use the following definition of coeffect scalars (for simplicity, we only
consider flat systems):

\begin{definition}
A \emph{\cclrd{flat coeffect lattice}} $(\C, \cseq, \cleq, \cunit, \czero)$ is a set
$\C$ together with elements $\cunit, \czero \in \C$ and an ordering $\cleq$ such
that $(\C, \cseq, \cunit)$ is a monoid and $(\C, \cleq)$ is a lattice with the least
element $\czero$.

Assuming $\cpar$ and $\czip$ are the \emph{least upper bound} and \emph{greatest lower bound}
of the lattice, we require the following distributivity axioms:
%
\begin{align*}
\quad (\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
\quad \cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
\end{align*}
\end{definition}

>>>>

The prototype implementation discussed in the previous chapter directly follows the two
coeffect calculi presented in this thesis. This links together all aspects of the coeffect
theory (type system, semantics, safety proofs) and it provides practical support for the
presented theoretical work. However, the theory of coeffects could be embedded in practical
programming languages in a various ways and the theory could be further developed in a number
of different directions.

The most important alternative directions revisit three design decisions made in this thesis.
First, rather than discussing flat and structural coeffects, we can treat them in a single
unified calculus tracking both kinds of context dependence. Second, rather than using ``language
semantics'' style (Section~\ref{sec:path-sem-langs}), we could use a ``meta-language'' style
and extend the host language with explicit constructs for working with coeffects. Third,
rather than implementing a prototype coeffect language, we could embed coeffects into an
existing language using techniques inspired by Haskell's ``do'' notation.

\paragraph{Chapter structure and contributions}
\begin{itemize}
\item We start by unifying the two kinds of coeffects discussed so far. We introduce the
  \emph{unified coeffect calculus} (Section~\ref{sec:further-unified}), which generalizes flat
  and structural systems and can be instantiated to track both flat and structural properties.

\item We discuss an alternative approach to defining coeffect systems based on the meta-language
  style (Section~\ref{sec:further-meta}), which highlights the relationship between our work and
  related work arising from modal logics such as CMTT \cite{logic-cmtt}.

\item We consider a way of implementing coeffects based on embedding comonadic computations
  via a lightweight syntactic extension and a rich type systems akin to Haskell's ``do'' notation
  (Section~\ref{sec:further-impl}).

\item Finally, we discuss extensions needed for real-world source languages with constructs
  such as conditionals and further applications of coeffects arising from substructural and bunched
  logics (Section~\ref{sec:further-theory})
\end{itemize}

@@

\begin{itemize}
\item The structural coeffect calculus contains explicit rules for context manipulation
  (weakening, contraction, exchange). In the flat coeffect calculus, these rules are not defined
  explicitly, but they can be derived thanks to the fact that variables in the context can be
  freely reordered.

@@

\item In the flat coeffect calculus, we distinguish between \emph{splitting} context requirements
  and \emph{merging} of context requirements ($\cpar$ and $\czip$, respectively). In the structural
  coeffect calculus, the operations (which model appending and splitting vectors) are invertible
  and so the structural coeffect algebra uses the same tensor product $\atimes$. However, the
  $\cpar$ operation is still needed in the contraction rule.
\end{itemize}

\noindent
In the unified calculus presented in this section, we address the three differences as follows.
We include explicit rules for context manipulation in the calculus; in systems that arise from flat
calculi, the structural rules do not change coeffects and can thus be applied freely. We generalize
the structure of coeffect annotations using the notion of a ``container'' which can be specialized
to obtain a single annotation or a vector of annotations. This could potentially be used to capture
other structures such as trees in bunched typing \cite{substruct-bunched}. Finally, we distinguish
between splitting and merging of context requirements (using the notation $\apar$ and $\azip$,
respectively). For structural coeffect calculi, the two operators coincide, but for flat
coeffect calculi, they differ and provide the needed flexibility.

\subsection{Shapes and containers}
Our generalization of coeffect structure using a \emph{coeffect container} is based on containers of
Abbott et al. \cite{types-containers}. Containers have later been linked to comonads by Ahman et al.
\cite{comonads-containers}, but here we use them as part of the coeffect algebra, rather than to
provide semantics of context-aware languages.

Intuitively a container describes data
types such as lists, trees or streams. A container is formed by shapes (\eg~lengths of
lists)\footnote{Shapes can also be intuitively thought of as sizes, but this is not fully precise
as some containers may have multiple shapes of the same length, \eg~differently balanced trees.}.
For every shape, we can obtain a set of positions in the container (\eg~offsets in a list of a
specified length). More formally:

@@

\begin{definition}
A \emph{\cclrd{unified coeffect scalar}} $(\C, \,\cseq, \,\cpar, \,\cunit, \,\czero, \,\cleq)$ is a set
$\C$ together with elements $\cunit, \,\czero \in \C$, binary operations $\cseq, \,\cpar$ such that
$(\C, \,\cseq, \,\cunit)$ and $(\C, \,\cpar, \,\czero)$ are monoids and a binary relation $\cleq$ such
that $(\C, \cleq)$ is a pre-order. That is, for all $r,s,t\in \C$:
%
\begin{equation*}
\begin{array}{ccr}
r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t  &
\cunit \;\cseq\; r = r = r \;\cseq\; \cunit &
\textnormal{(monoid)}
\\
r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
\czero \;\cpar\; r = r = r \;\cpar\; \czero &
\textnormal{(monoid)}
\\
\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t&
t\; \cleq\; t &
\textnormal{(pre-order)}
\\
\end{array}
\end{equation*}
%
The following distributivity axioms hold:
\begin{align*}
\quad (\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
\quad \cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
\end{align*}
%
In addition, we require the following two properties of $\czero, \cpar$ and $\cleq$:
\begin{equation*}
\begin{array}{l}
 \quad \cclrd{r}\,\cleq\,\cclrd{r'} \; \Rightarrow \; \forall\cclrd{s}.\;(\cclrd{r}\,\cseq\,\cclrd{s})\;\cleq\;(\cclrd{r'}\,\cseq\,\cclrd{s}) \\
 \quad \czero\;\cleq\;(\czero\,\cseq\,\cclrd{r})
\end{array}
\end{equation*}
\end{definition}

@@

In the coeffect judgment $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$, the coeffect annotation
$\aclrd{\textbf{r}}$ is drawn from the set of coeffect scalars $\C$ indexed by the shape corresponding
to $\Gamma$. We write $\sclrd{s} = \slen{\Gamma}$ for the shape corresponding to $\Gamma$. The operation
$\SP(\sclrd{s})$ returns a \emph{set} of positions and so we can write $\aclrd{\textbf{r}} \in
  \SP(\sclrd{s}) \rightarrow \C$  as a mapping from positions (defined by the shape) to scalar coeffects.
We write this as the exponent $\aclrd{\textbf{r}} \in \C^{\SP(\sclrd{s})}$.
The coeffect container is equipped with an operation that appends shapes (when concatenating
variable contexts) and two special shapes in $\SH$ representing the empty context and the singleton
context.

@@

Therefore, a coeffect annotation is a vector
$\aclrd{\textbf{r}} \in \C^{\SP(\sclrd{n})}$ and assigns a coeffect scalar $\aclrd{\textbf{r}}(i) \in \C$
for each position (corresponding to a variable $x_i$ in the context).
\end{example}

@@

\begin{example}
Similarly, we can also define a coeffect container with \emph{no} positions,
\ie~$\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$ where $\SH = \{\sflat \}$, $\SP(\sflat) = \emptyset$,
${|}\Gamma{|}=\sflat$ and $\sflat\,\diamond\,\sflat = \sflat$.
This reduces our system to the simply-typed $\lambda$-calculus with no context annotations, because
$\SP(\sflat)=\emptyset$ and so coeffect annotations would be from the empty set $\C^\emptyset$.
\end{example}

@@

\begin{definition}
Given a \cclrd{unified coeffect scalar} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a
\sclrd{coeffect container} $(\SHP, \slen{-}, \stimes, \sempty, \sunit)$ a \aclrd{unified
coeffect algebra} extends the two structures with $(\azip, \apar, \azero)$ where $\azero \in \C^{\SP(\sempty)}$
is a coeffect annotation for the empty context and $\azip, \apar$ are families of operations that
combine coeffect annotations indexed by shapes. That is $\forall \sclrd{n},\sclrd{m} \in \SH$:

\begin{equation*}
\apar_{\sclrd{m},\sclrd{n}}, \azip_{\sclrd{m},\sclrd{n}} ~:~ \C^{\SP \sclrd{m}} \times \C^{\SP \sclrd{n}} \rightarrow \C^{\SP (\sclrd{m}\,\stimes\,\sclrd{n})}
\end{equation*}
\end{definition}

\noindent
A coeffect algebra induces the following three additional operations:
%
\begin{equation*}
\begin{array}{rclcrcl}
 \alift{-} &\narrow{:}& \C \rightarrow \C^{\SP(\sunit)}  \\
 \alift{\cclrd{x}} &\narrow{=}& \lambda \_ . \cclrd{x}   \\
\end{array}
\begin{array}{rclcrcl}
 \aseq_{\sclrd{m}} &\narrow{:}&  \C \times \C^{\SP(\sclrd{m})} \rightarrow \C^{\SP(\sclrd{m})}   \\
 \cclrd{r}\, \aseq\, \aclrd{\textbf{s}} &\narrow{=}& \lambda \sclrd{s} . \cclrd{r}\, \cseq\, \aclrd{\textbf{s}}(\sclrd{s}) \\
\end{array}
\begin{array}{rclcrcl}
 \alen{-} &\narrow{:}& \C^{\SP(\sclrd{m})} \rightarrow \sclrd{m}\\
 \alen{\aclrd{\textbf{r}}} &\narrow{=}& \sclrd{m}
\end{array}
\end{equation*}
%
The $\alift{-}$ operation lifts a scalar coeffect to a shape-indexed coeffect that is indexed by
the shape of a singleton context. The $\aseq_{\sclrd{m}}$ operation is a left multiplication of a
vector by a scalar. As we always use bold face for vectors and ordinary face for scalars (as well
as a distinct colour), using the same symbol is not ambiguous. We also tend to omit the subscript
$\sclrd{m}$ and write~$\aseq$.

@@

The type system for the unified coeffect calculus is parameterized by a coeffect scalar
$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ together with a coeffect algebra $(\azip, \apar, \azero)$
and the derived constructs $\alift{-}$, $\alen{-}$ and $\aseq$.
As in the structural system, free-variable contexts $\Gamma$ are treated as vectors (without duplicate
variables) with associativity built-in. The order of variables matters, but can be changed
using the exchange rule. The context annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$
are shape-indexed coeffects (rather than simple vectors as before). As before, functions are
annotated with coeffects scalars.

@@

Note that we use the terms \emph{merging} and \emph{splitting} in the syntactic (top-down) sense. As
discussed in Section~\ref{sec:flat-calculus-understanding}, we can also read
the rules in semantic (bottom-up) sense, in which case assumptions merge available contextual
information and conclusions split available contextual information.

@@
